Multi-level modelling with Openflexo/FML - A contribution to the \mpc

### 2021-05-25 4-Technology.tex
Updates from Overleaf

To meet the \mpc, we have decided to use the general
approach of model federation~\cite{Golra2016-federation}. \emph{Model
  federation} is a way models are assembled together with some kind of
low-coupling links. It has been studied to answer the OMG' RFP
Semantic Modeling for Information Federation (SIMF)~\cite{simf}. In
contrast to approaches that compose meta-models into a single large
meta-model welcoming all needed entities, model federation build links
among models and meta-models (even through levels) to make ``things''
work together. As an illustration of this feature, we developed a
free-modeling editor -- freeing oneself from the bonds of
model/meta-model conformity -- that is presented
in~\cite{models2016-freemodel}. Another very interesting feature of
this approach is the strong decoupling among tools that remain usable
after federations are made.

@@

This framework relies on the following architecture. A federation
gathers a set of conceptual models, named \emph{virtual models} and a
set of \emph{federated models}. Each federated model pertains to a
\emph{technological space} and uses the language of its specific
paradigm while a virtual model is built using the Flexo Modeling
Language (FML). Each federated model can be viewed as an autonomous
component that may evolve with its own tooling. The virtual models
serve as control components.
% Figure~\ref{fig:space-ex} presents a
% simplified version of the \textsf{Goal-xls} federation example that
% groups these different models. The upper half of the figure
% illustrates the design of a federation with a model coming from the
% \smk technological space (the goal diagram), one from the Excel
% technological space (the file) and one virtual model, named
% \emph{ReqXls}, reifying the dependencies between the goal model and
% the reporting file.

A \emph{technology adapter} (TA) is a reusable library that defines
connections between the FML execution engine and a particular
technological space. The model federation framework provides ways to
define TA.
%The \emph{ReqXls} virtual model relies on two technology
%adapters (Excel TA and \smk TA) for manipulating the models in their
%respective technological spaces.

FML is a language designed to define virtual models. A virtual model
is composed of a set of \emph{concepts}, while itself being a concept.
Hence, virtual models are structuring units while concepts are the
core entities. A concept has a set of \emph{roles} and
\emph{behaviors}. A parallel to object-oriented approach can be useful
to understand FML\footnote{Beware, even though useful for
  comprehension, this correspondence is not
  reliable, as some aspects of FML do not map to object oriented
  concepts}. A concept corresponds to a class, its roles to the
attributes of the class and its behaviors to the methods of the class.
% In our example, a concept \texttt{Requirement} is presented with
% three roles \texttt{row}, \texttt{goal} and \texttt{refines}. The
% first two are represented as attributes whereas the third as an
% association.
These roles have types defining the kind of value the role will point
at runtime.
% Our example illustrates three forms of such types \ie another
% concept for \textsf{refines}, a type defined in a TA for
% \textsf{Row} or a type defined in an external model for
% \textsf{Goal}.
Whenever a type external to the federation space (from a TA or an
external model) is used, one needs to use a \emph{model slot}. A model
slot is a mediation entity, associated with a TA, in charge of giving
access to external elements of the corresponding technological space.

@@

The tool support for model federation framework,
Openflexo\footnote{\url{https://github.com/openflexo-team}}, is
developed as an open source initiative. This tool offers a FML
execution engine with an interactive virtual model design environment.
It has been used in several case studies including model mapping,
multi-paradigm process modeling and enterprise architecting. It has
also been used to build a tool, the freemodeling editor that has been
used in industrial projects~\cite{models2016-freemodel}. As of today, this tool
offers some mature technology adapters (\emph{docx} and \emph{excel}
for documents, EMF and OWL for modeling languages, JDBC for databases,
REST and XMLRPC for external services and one for diagraming tools)
and other more rudimentary (pdf, HTTP, XML, etc.).

>>>>

\noteFabien{metamodel ou meta-model ? perso je penche plus pour le premier}
\noteJC{pareil, j'utilise plutôt metamodel, en général}

To meet the \mpc, we have decided to use the general
approach of model federation~\cite{Golra2016-federation}. \emph{Model
  federation} is a way to assemble models using some kind of
low-coupling links. It has been studied to answer the OMG' RFP
Semantic Modeling for Information Federation (SIMF)~\cite{simf}. In
contrast to approaches that compose meta-models into a single large
meta-model grouping all needed entities, model federation build links
among models and meta-models (even through levels) to make ``things''
work together. As an illustration of this feature, we developed a
free-modeling editor -- freeing oneself from the bonds of
model/meta-model conformity -- that is presented
in~\cite{models2016-freemodel}. Another very interesting feature of
this approach is the strong decoupling among tools that remain usable
after federations are made.

@@

This framework relies on the following architecture. A federation
gathers a set of conceptual models, named \emph{virtual models} and a
set of \emph{federated models}. Each federated model pertains to a
\emph{technological space} and uses the language of its specific
paradigm while a virtual model is built using the Flexo Modeling
Language (FML). Each federated model can be viewed as an autonomous
component that may evolve with its own tooling. The virtual models
serve as control components.

FML is a language designed to define virtual models. A virtual model
is composed of a set of \emph{concepts}, while itself being a concept.
Hence, virtual models are structuring units while concepts are the
core entities. A concept has a set of \emph{roles} and
\emph{behaviors}. A parallel to object-oriented approach can be useful
to understand FML\footnote{Beware, even though useful for
  comprehension, this correspondence is not
  reliable, as some aspects of FML do not map to object oriented
  concepts}. A concept corresponds to a class, its roles to the
attributes of the class and its behaviors to the methods of the class.
These roles have types defining the kind of value the role will point
at runtime.
Whenever a type external to the federation space (from a TA or an
external model) is used, one needs to use a \emph{model slot}. A model
slot is a mediation entity, associated with a TA, in charge of giving
access to external elements of the corresponding technological space.

A \emph{technology adapter} (TA) is a reusable library that defines
connections between the FML execution engine and a particular
technological space. The model federation framework provides ways to
define TA.

@@

The tool support for model federation framework,
Openflexo\footnote{\url{https://github.com/openflexo-team}}, is
developed as an open source initiative. This tool offers a FML
execution engine with an interactive virtual model design environment.
It has been used in several case studies including model mapping,
multi-paradigm process modeling and enterprise architecting. It has
also been used to build a tool, the freemodeling editor that has been
used in industrial projects~\cite{models2016-freemodel}. As of today, this tool
offers some mature technology adapters (\emph{docx} and \emph{excel}
for documents, EMF and OWL for modeling languages, JDBC for databases,
REST and XMLRPC for external services and one for diagramming tools)
and other more rudimentary (pdf, HTTP, XML, etc.).

### 2021-05-25 6-ModelPresentation.tex
Updates from Overleaf

\textit{Code} concept extends \textit{SEArtifact}, and defines a relation \textit{language} to \textit{ProgrammingLanguage} with single cardinality (\textbf{S4}). \textbf{S5} is more ambigous as task type \textit{Coding} defines one or more programming languages but produces code which is expressed in one language only. This requirement is captured through the redefinition of \textit{declaresRequiredProducedArtifacts()} where programming language should also match \noteSylvain{Pas la place pour mettre le code, non nécessaire si ?}. \textbf{S6} is guaranteed though the \textit{language} relation defined in \textit{Code} and following invariant declared in \textit{Code}:

\begin{lstlisting}[breaklines=true, language=java, basicstyle=\ttfamily\scriptsize, mathescape=true]
forEach (artifactType : types) {
    assert !(artifactType instanceof CodeArtifactType) | artifactType.isSenior
}
\end{lstlisting}

%

Code artifact must be produced for each programming language defined in a \textit{CodingTaskType}. If \textit{COBOL} is referenced in languages defined for a given Coding instance, then we must instantiate a \textit{Code} instance with \textit{COBOL} language (\textbf{S5} and \textbf{S6}).

\begin{lstlisting}[breaklines=true, language=java, basicstyle=\ttfamily\scriptsize, mathescape=true]
concept CodingTaskType extends SETaskType 
{
  ...
  public Artifact[] instantiateArtifacts() {
    Artifact[] returned = {}; 
    for (langage : languages) {
      returned.add(new Code(langage); 
    }
    return returned;
  }
  ...
}    
\end{lstlisting}

>>>>

\textit{Code} concept extends \textit{SEArtifact}, and defines a relation \textit{language} to \textit{ProgrammingLanguage} with single cardinality (\textbf{S4}). \textbf{S5} is more ambigous as task type \textit{Coding} defines one or more programming languages but produces code which is expressed in one language only. This requirement is captured through the redefinition of \textit{declaresRequiredProducedArtifacts()} where programming language should also match. \noteSylvain{Pas la place pour mettre le code, non nécessaire si ?} \textbf{S6} is guaranteed though the \textit{language} relation defined in \textit{Code} and following invariant declared in \textit{Code}:

\begin{lstlisting}[breaklines=true, language=java, basicstyle=\ttfamily\scriptsize, mathescape=true]
forEach (artifactType : types) {
    assert !(artifactType instanceof CodeArtifactType) | artifactType.implements(languages)
}
\end{lstlisting}

\textit{Ann Smith} is the only one allowed to perform coding in COBOL (\textbf{S7}). This is implemented through the redefinition of \textit{newTask(String)} in \textit{CodingTaskType}:

\begin{lstlisting}[breaklines=true, language=java, basicstyle=\ttfamily\scriptsize, mathescape=true]
concept CodingTaskType extends SETaskType 
{
  ...
  public Coding newTask(String name) {
    Coding returned = new Coding(name,this); 
    if (languages.contains(ProgrammingLanguage.COBOL)) {
        returned.addToPerformingActors(getActor("Ann Smith"));
    }
    return returned;
  }
}    
\end{lstlisting}

\textbf{S8} requirement is simply captured with the definition of \textit{Testing} instance of \textit{SETaskType}, \textit{Tester} instance of \textit{ActorType}, and \textit{Test report} instance of \textit{ArtifactType}.
% TODO S9

All software engineering artifacts defined in the context of Acme Software Engineering Process (figure \ref{fig:AcmeSoftwareDevelopmentProcess}) are all instances of \textit{SEArtifact} which defines both attributes \textit{responsible} (an \textit{Actor} instance), and \textit{versionNumber} (an \textit{integer} value), and thus fulfills \textbf{S10}. \textit{Bob Brown} is declared as an instance of \textit{Actor}, and references \textit{Analyst} and \textit{Tester} (\textit{ActorType} instances). He is also referenced by all instances of \textit{SETaskType} as the creator for related task types (\textbf{S11}).

\textbf{S12} is somehow ambigous as it implies an implicit semantics regarding underlying business logic of process execution. We assume that all tasks may define an expected duration, which might be checked during process execution. This can be modelled through the addition of \textit{expectedDuration} attribute in \textit{TaskType}. Alternatively, this can be modelled through the definition of \textit{SETesting} concept, as a specialization of \textit{SETaskType}, and the instantiation of \textit{Testing} as an instance of \textit{SETesting}. The business logic expressed by \textbf{S12} requirement should be redefined in \textit{SETesting}.

\textbf{S13} requirement has been previously partially fulfilled. This must be completed with...

% TODO S13

### 2021-05-25 7-RequirementsSatisfaction.tex
Updates from Overleaf

>>>>

\noteJC{en fait, la section précédente fait déjà le travail, donc ça fait effectivement redandant. Dans ce cas, pourquoi le répéter ? On peut éventuellement fusionner les deux sections. L'appel impose plus ou moins le plan, mais tant que les différents ponits sont présents, ça ne doit pas trop déranger. Ton histoire de tableau pourrait être une sorte de résumé de la section précédente.}
